Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees

### 2023-09-12 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Paper updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3278 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e88d07fc867a04fa9afe5e1b7b944993f259565e

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope %(presented in Table~\ref{tab:local_coefs}). 
Groups of children with lower (higher) SES also have a lower (higher) intercept, indicating lower (higher) average science ability. For children with lower SES (node~2), there is also an effect of gross motor skills: The group with lower motor skills has a lower intercept. For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.

@@

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
~
\begin{subfigure}{0.25\textwidth}%
\vspace{-2.95cm}
\hspace{-2.5cm}
\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
\label{subfig:coefs}
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The table presents numerical estimates of fixed intercepts and slopes.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

In terms of MAD, however, LMM trees initializing estimation with the random effects performed best, but only if the random slope was not estimated. Initializing estimation with the random effects may only be useful if the random-effects specification is kept relatively simple. With more complex random-effects specifications, initializing estimation seems to result in too few groups being detected; subgroup differences are then more likely to be captured by the random effects than by the tree structure. The second-lowest MAD was observed for LM trees with clustered covariances (left panel). Combined use of random-effects initialization and cluster-level covariances was not very effective, irrespective of the random-effects specification.

Distributions of the number of splits, separated according to the levels of the data-generating design are presented in Figure~\ref{fig:LMM_sizes_interact} and discussed in Appendix~\ref{sec:AppendixA}. The results show a pattern very similar to Figure~\ref{fig:LMM_sizes}; no substantial interactions between data-generating and model-specification parameters were observed. Main effects of the data-generating parameters were as expected: The strongest effects were for $\sigma_{b_1}$ and $N$, with higher values resulting in a higher number of splits. The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, while the correlation between partitioning variables hardly affected the number of implemented splits.

Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split: Only very rarely is $u_1$ not selected for the first split.

@@

We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and the approaches that performed well in the simulations:
% 
\begin{itemize}
  \item An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests.
  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and initialization assuming zero random effects.
  \item An LMM tree using clustered covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and initialization assuming zero random effects.
  \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$) and initialization assuming zero random effects . 
  \item An LMM tree uwith a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effect initialization with the full-sample estimate. 
\end{itemize}
%
Although LRT-based SEM trees performed very well in the simulations, they could not be used in this study because growth curve SEMs do not allow for incorporating continuous time.

@@

Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions. Differences in predictive performance are small, all $R^2$ differences are smaller than 0.01. Bonferroni-adjusted pairwise $t$-tests indicated no difference in performance between default LMM trees and LMM trees with random-effects initialization for any of the three outcomes. Incontract, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes. For the science outcomes, no significant differences were observed. Thus, cluster-level covariances seem to provide the most robust improvement of performance.

Figure~\ref{fig:application_sizes} presents tree size distributions. Similar to the simulation study's results, default LMM trees implement the largest number of splits. Cluster-level covariances provide a robust reduction in the number of splits. Given their non-distinguishable predictive performance, LM trees with clustered covariances may be preferred if a obtaining a sparse result is critical. LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes have been specified. Random-effects initialization yields tree sizes similar to the default fitting approach.

@@

Strong performance of clustered covariances was also observed in partitioning real-world academic trajectories. They provided substantially smaller trees for all outcomes and better or equal predictive accuracy. In comparison to cluster-level covariances, random-effects initialization resulted larger trees for all outcomes and worse performance for the reading and math outcomes.

@@

The current evaluations were limited to Gaussian responses and LGCMs. Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses. Also, whether our conclusions generalize to settings beyond LGCMs remains to be evaluated. We expect that the strong performance of cluster-level covariances in the parameter stability tests generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures. Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances. Though computationally more burdensome, future work could assess potential benefits of using the full sandwich estimator.

>>>>

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope. Groups of children with higher SES also have higher intercepts, indicating higher average science ability. The group of children with lower SES (node~2) is further split based on gross motor skills, with higher motor skills resulting in a higher intercept. The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.

@@

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
~
\begin{subfigure}{0.25\textwidth}%
\vspace{-2.95cm}
\hspace{-2.5cm}
\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
\label{subfig:coefs}
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories. The table presents numerical estimates of fixed intercepts and slopes.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

In terms of MAD, however, LMM trees initializing estimation with the random effects performed best, but only if the random slope was not estimated. Thus, initializing estimation with the random effects may only be useful if the random-effects specification is kept relatively simple. With more complex random-effects specifications, actual subgroup differences may be more likely captured by the random effects than by the tree structure. The second-lowest MAD was observed for LM trees with clustered covariances (left panel). Combined use of random-effects initialization and cluster-level covariances was not very effective, irrespective of the random-effects specification.

Distributions of the number of splits, separated according to the levels of the data-generating design are presented in Figure~\ref{fig:LMM_sizes_interact} and discussed in Appendix~\ref{sec:AppendixA}. The results show a pattern very similar to Figure~\ref{fig:LMM_sizes}; no substantial interactions between data-generating and model-specification parameters were observed. Main effects of the data-generating parameters were as expected: Strongest effects were for $\sigma_{b_1}$ and $N$, with higher values resulting in a higher number of splits. The number of noise variables and $\sigma^2_{b_1}$ had smaller effects, while the correlation between partitioning variables hardly affected the number of implemented splits.

Table~\ref{tab:first_splits} shows the variables selected for the first split, and indicates high accuracy for recovery of the first split for all LM(M) tree fitting approaches. Only very rarely is $u_1$ not selected for the first split, if a first split was implemented.

@@

We applied five LM(M) trees to the data, focusing on the original GLMM tree approach and approaches that performed well in the simulations:
% 
\begin{itemize}
  \item An LM tree ($\hat{\sigma}_{b_0} = \hat{\sigma}_{b_1} = 0$) using clustered covariances in the parameter-stability tests.
  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and initialization assuming zero random effects.
  \item An LMM tree using clustered covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and initialization assuming zero random effects.
  \item An LMM tree using clustered covariances, with both random intercept and slope freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} > 0$) and initialization assuming zero random effects . 
  \item An LMM tree using observation-level covariances, with a random intercept freely estimated ($\hat{\sigma}_{b_0} > 0$, $\hat{\sigma}_{b_1} = 0$) and random-effect initialization with the full-sample estimate. 
\end{itemize}
%
Although LRT- and score-based SEM trees performed very well in the simulations, they could not be used in this study because growth-curve SEMs do not allow for incorporating continuous time.

@@

Figure~\ref{fig:application_MSEs} and Table~\ref{tab:application_MSEs} present MSE distributions. Differences in predictive performance are small, all $R^2$ differences are smaller than 0.01. Bonferroni-adjusted pairwise $t$-tests indicated no difference in performance between default LMM trees and LMM trees with random-effects initialization for any of the three outcomes. In contrast, the three LM(M) trees using clustered covariances performed significantly better for the math and reading outcomes. For the science outcomes, no significant differences were observed. Thus, cluster-level covariances seem to provide the most robust improvement of performance.

Figure~\ref{fig:application_sizes} presents tree size distributions. Similar to the simulation study's results, default LMM trees implement the largest number of splits. Cluster-level covariances provide a robust reduction in the number of splits. Given their non-distinguishable predictive performance, LM trees with clustered covariances may be preferred if a obtaining a sparse result is critical. LMM trees with clustered covariances may provide less sparsity but better predictive accuracy, irrespective of whether random intercepts and/or slopes were estimated. In contrast to the simulation study results, random-effects initialization here yields tree sizes similar to the default fitting approach.

@@

Strong performance of clustered covariances was also observed in partitioning real-world academic trajectories. They provided substantially smaller trees for all outcomes and better or equal predictive accuracy. In comparison to cluster-level covariances, random-effects initialization resulted in larger trees for all outcomes and worse performance for the reading and math outcomes.

@@

The current evaluations were limited to Gaussian responses and LGCMs. Future studies should assess performance of GLMM trees in partitioning longitudinal data with, for example, binomial or count responses. We expect that the strong performance of cluster-level covariances generalizes to other settings where covariates are measured at higher levels, either in longitudinal and/or otherwise nested data structures, but whether our conclusions generalize to settings beyond LGCMs remains to be evaluated. Finally, we used the outer-product-of-gradients (OPG) estimator for computing (clustered) covariances. Though computationally more burdensome, future work could assess potential benefits of using the full sandwich estimator.
